Performance of Individual Audit Filters in Predicting Opportunities for Improvement

### 2024-06-10 manuscript.Rmd
update

---
title: "Performance of Individual Audit Filters in Predicting Opportunities for Improvement
  in Adult Trauma Patients"
subtitle: A single-center registry-based retrospective cohort
output: html_document
author: "Josef Al-Khalili"
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

Abstract
========
*Introduction:* Trauma is responsible for 9% of global deaths annually
and has a significant impact on individuals and society. Quality
Improvement (QI) programs seek to minimize these adverse outcomes using
Audit Filters (AFs) and interdisciplinary Morbidity and Mortality
Conferences (MMC). AFs are based on criteria for standard care and can
function as a screening tool of eligible patient cases for discussion at
MMCs. One of the goals with MMCs is to determine potential Opportunities
for Improvement (OFI). *Aims:* This study aimed to evaluate the
performance of individual AFs in predicting OFIs and to assess their
relevance as a screening tool for Morbidity and Mortality Conferences
(MMC). *Material and Methods:* Register data from 8,309 patient cases
were collected for analysis. The sensitivity, specificity and Area Under
the Receiver Operating Characteristics Curve (AUC) was calculated to
assess performance. An empirical bootstrapping test of AUC values was
conducted to compare the performance of each AF to coincidence (AUC =
0.5) and minimum AUC for diagnostic tests (AUC = 0.8). *Results:* Four
out of the 11 AFs assessed had an AUC of 0.5 (p \> 0.05). The remaining
AFs demonstrated a correlation with OFI, with AUC ranging from 0.51 to
0.63. The AF "\> 60 min until first intervention" exhibited the highest
AUC (0.63, 95% CI [0.61-0.65]). *Conclusion:* No individual AF seemed to
perform well in predicting OFI, which suggest rethinking the current
screening method for MMCs.

Background
----------

Methods
-------

Results
-------

Conclusion
----------

Introduction
============
Trauma, defined as a physical injury and
the body's associated response, accounts for 8-9%
of global deaths annually [@world2014injuries; @roth2018global], mainly affecting younger individuals in low- and middle-income countries attributed to mainly road traffic injuries [@vos2020global;
@roth2018global; @SweTrau2022Arsrapport]. Furthermore, trauma is responsible for the top cause for disability-adjusted life-years (DALY) [@vos2020global] and a higher propensity towards mental disorders and cognitive decline [@kessler1995posttraumatic;
@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts; @visser2017course]. This negatively affects individuals, communities and society as a whole [@newnam2014impacts].It is estimated that the cost of road traffic injuries alone
accounts for 2% of gross domestic product (GDP) in high-income countries
[@world2014injuries].

@@

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->

Methods 
=======

Study design
------------

@@

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->

Variables and data sources/measurements
---------------------------------------

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

Bias 
----

@@

Study size 
----------

@@

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------

<!--

@@

(e) Describe any sensitivity analyses
-->

Results
=======

@@

Discussion
==========
This study showed that AFs do not perform well in predicting OFIs. 
AFs generally exhibited a high specificity relative to their sensitivity
which may be attributed to the outcome being rather uncommon, whilst
every individual AF misses the majority of OFIs. This could suggest that
the AFs are too rigid in their criteria. On the contrary, a too
inclusive criteria might flag several patients, who do not necessarily
have an OFI [@willis2008interpreting].

AFs are static in their criteria, contrary to health care in general.
Consequently, the utility of AFs may diminish if confronted with
emerging challenges or upon resolution of the quality gaps they were
designed for. This highlights their need for
consistent and evidence-based review to determine current usefulness
[@santana2014development]. However, such
review process might present with costs greater than the potential
benefit from AFs.

Higher ISS, longer times to CT, longer times to first intervention and higher levels of care were correlated with OFIs, corresponding with a previous study conducted at Karolinska University Hospital by Albaaj
et al. [@albaaj2023patient]. Greater severity and higher levels of care call for greater complexity, requiring more interventions, allowing for more mistakes. This could explain their correlation with OFI[@albaaj2023patient].

Notably, GCS at ED showed a correlation with OFI even though the median
GCS and its IQR was identical for individuals with OFI and without. Upon
further analysis this could be explained by the difference in
distribution outside the IQR.

The AF "GCS<9 and not intubated" showed no significant ability to predict OFIs, similarly shown by Willis et al. [@willis2008interpreting] in 2008. An
explanation for this could be that an abnormal GCS as a result from
intoxication versus head injury for example are not comparable, suggesting that the AF might be more reflective of clinical
decision rather than quality of care [@willis2008interpreting]. However, according to Sanddal et al. [@sanddal2011analysis] airway management was
associated with OFI, suggesting that an alternative AF related
to airway management might perform differently.

Both AFs measuring delay: ">60 until first intervention" and ">30min
until first CT", showed a correlation with OFI. This is comparable to
what Teixeira et al. [@teixeira2007preventable] demonstrated, with delay
correlating to preventable death. Furthermore, Ghorbani et al.
[@ghorbani2018analysis] identified delay in treatment and delay to CT as common errors in trauma care. A more recent
study in Japan by Yamamoto et. al [@yamamoto2023immediate] showed a
significant reduction in mortality when patients received a whole-body
CT within 10 minutes of arrival at the ED. Notably,
OFI in this study was not soley a reflection of mortality but also morbidity. Rather, it
highlights the difference between hospitals, which AFs are in use, and
how they are assessed.

AFs that present with a high risk of inevitable death: "Dead
at 30 days" and "CPR and thoracotomy" showed no correlation with OFI. This could be explained by inevitable
mortality, thus presenting with a smaller time frame for mistakes to be
made.

@@

@@

AFs are widely used but generalizability is complicated due to the lack
of common grounds on what they are, what purpose they serve and how they
shall be used [@stelfox2010quality; @stelfox2011evidence;
@berg2022perceived].

Our findings suggest that AFs perform poorly in predicting an OFI. The AFs
with performance exceeding coincidence (AUC = 0.5) might be a beneficial
addition to an already existing QI programme or
be used to fine-tune AFs already in use to plug hospitals' quality gaps.
Due to the study being conducted in a single, trauma level I centre in
Stockholm it is unrealistic to assume applicability of these findings at
other hospitals facing unique challenges
[@world2009guidelines]. For example, Low and middle income countries
seem to benefit from certain AFs that high income countries do not and
vice versa [@berg2022perceived].

## Strengths

As stated in the method section, some AFs were manually created from
registry data instead of relying on the selection process by the
specialised nurse (see table 1). This, in order to minimise
misclassification bias.

Furthermore, there has been very limited high quality evidence
supporting the efficacy of AFs in improving trauma care
[@evans2009audit; @stelfox2010quality] and predicting OFI
[@sanddal2011analysis; @roy2017learning]. This study shows comparable
results. Previous studies also focus primarily on AFs' ability to reduce
mortality [@chadbunchachai2001comparative; @chadbunchachai2003study]. In
this study the morbidity of patient cases is also taken into account
using OFIs as an umbrella term that includes factors ranging from bad
documentation to preventable death.

From 2017 all trauma patients admitted to Karolinska University Hospital
were included in the registry reducing the risk of bias in inclusion of
patient cases.

@@

With the ongoing academic, political, and public attention towards
patient safety and quality of healthcare [@evans2009audit], it is
important to discover optimal methods for monitoring and enhancing the
care provided within trauma systems.

Given the expanding field of artificial intelligence and machine
learning, newer studies should focus on using these tools to screen for
patient cases at risk of OFIs. This will diminish misclassification
rates and ensure consistent and sustained high performance, attributable
to an algorithm's capability to adapt and learn from the prevailing
context of trauma care. This approach is contemporary but not
unexplored. A preprint is currently available by Attergrim et al.
[@attergrim2023predicting] exploring the idea of machine learning models
as a screening tool for potential OFIs. They found that machine learning
models outperformed the AFs currently used at Karolinska University
Hospital in predicting OFIs.

@@

Conclusion
==========

AFs are widely used as a tool in quality improvement programmes and
should reflect the current gap in trauma care quality. Strong evidence
supporting their efficacy in quality improvement is scarce. In this
study, no audit filter seemed to show promising performance in
predicting OFIs with AUC ranging from 0.5 to 0.63. These results
highlight the importance of continually evaluating the relevance of AFs.
However, this process takes time and resources and using an alternative
to AFs is recommended.

References
==========

>>>>

---
title: "Performance of Individual Audit Filters in Predicting Opportunities for Improvement
  in Adult Trauma Patients"
subtitle: A single-center registry-based retrospective cohort
output: html_document
author: "Josef Al-Khalili"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

# Abstract

## Background

Trauma is responsible for 9% of global deaths annually
and has a significant impact on individuals and society. Quality
Improvement (QI) programs seek to minimize these adverse outcomes using
Audit Filters (AFs) and interdisciplinary Morbidity and Mortality
Conferences (MMC). AFs are based on criteria for standard care and can
function as a screening tool of eligible patient cases for discussion at
MMCs. One of the goals with MMCs is to determine potential Opportunities
for Improvement (OFI).This study aimed to evaluate the
performance of individual AFs in predicting OFIs and to assess their
relevance as a screening tool for Morbidity and Mortality Conferences
(MMC).

## Methods

Register data from 8,309 patient cases
were collected for analysis. The sensitivity, specificity and Area Under
the Receiver Operating Characteristics Curve (AUC) was calculated to
assess performance. An empirical bootstrapping test of AUC values was
conducted to compare the performance of each AF to coincidence (AUC =
0.5) and minimum AUC for diagnostic tests (AUC = 0.8).

## Results

Four
out of the 11 AFs assessed had an AUC of 0.5 (p \> 0.05). The remaining
AFs demonstrated a correlation with OFI, with AUC ranging from 0.51 to
0.63. The AF "\> 60 min until first intervention" exhibited the highest
AUC (0.63, 95% CI [0.61-0.65]).

## Conclusion

No individual AF seemed to
perform well in predicting OFI, which suggest rethinking the current
screening method for MMCs.

# Introduction

Trauma, defined as a physical injury and the body's associated response,
accounts for 8-9% of global deaths annually [@world2014injuries;
@roth2018global], mainly affecting younger individuals in low- and
middle-income countries attributed to mainly road traffic injuries
[@vos2020global; @roth2018global; @SweTrau2022Arsrapport]. Furthermore,
trauma is responsible for the top cause for disability-adjusted
life-years (DALY) [@vos2020global] and a higher propensity towards
mental disorders and cognitive decline [@kessler1995posttraumatic;
@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts;
@visser2017course]. This negatively affects individuals, communities and
society as a whole [@newnam2014impacts].It is estimated that the cost of
road traffic injuries alone accounts for 2% of gross domestic product
(GDP) in high-income countries [@world2014injuries].

@@

```{=html}
<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->
```
# Methods

## Study design

@@

## Setting

```{=html}
<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
```
## Participants

```{=html}
<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
```
```{=html}
<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->
```
```{=html}
<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
```
## Variables and data sources/measurements

```{=html}
<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
```
## Bias

@@

## Study size

@@

## Quantitative variables

```{=html}
<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->
```
## Statistical methods

```{=html}
<!--

@@

(e) Describe any sensitivity analyses
-->
```
# Results

@@

# Discussion

The aim of this study was to evaluate the performance of individual AFs in predicting OFIs and 
to assess their suitability as a screening tool for MMCs. This study showed that no AFs was significantly successful in predicting OFIs. Four AFs ("SBP<90," "dead at 30 days," "GCS<9 and not intubated," and "CPR and thoracotomy") were equivalent to random chance in their predictive ability. The highest AUC was for ">60 minutes until first intervention,". Notably, all AFs performed significantly below the accepted standard for diagnostic tests (>0.8).

@@

Higher ISS, longer times to CT, longer times to first intervention and
higher levels of care were correlated with OFIs, corresponding with a
recent study conducted at Karolinska University Hospital by Albaaj et
al. [@albaaj2023patient]. Notably, GCS at ED showed a correlation with OFI even though the median
GCS and its IQR was identical for individuals with OFI and without. Upon
further analysis this could be explained by the difference in
distribution outside the IQR.

The AF "GCS\<9 and not intubated" showed no significant ability to
predict OFIs, similarly shown by Willis et al. [@willis2008interpreting]
in 2008. An explanation for this could be that an abnormal GCS as a
result from intoxication versus head injury for example are not
comparable, suggesting that the AF might be more reflective of clinical
decision rather than quality of care [@willis2008interpreting]. However,
according to Sanddal et al. [@sanddal2011analysis] airway management was
associated with OFI, suggesting that an alternative AF related to airway
management might perform differently.

Both AFs measuring delay: "\>60 until first intervention" and "\>30min
until first CT", showed a correlation with OFI. Comparably, Teixeira et al. [@teixeira2007preventable] found a correlation between delay
and preventable death. Furthermore, Ghorbani et al.
[@ghorbani2018analysis] identified delay in treatment and delay to CT as
common errors in trauma care. A more recent study in Japan by Yamamoto
et. al [@yamamoto2023immediate] showed a significant reduction in
mortality when patients received a whole-body CT within 10 minutes of
arrival at the ED. Notably, OFI in this study was not soley a reflection
of mortality but also morbidity. Rather, it could highlight differences in quality gaps
between hospitals.

"Dead at 30 days"
and "CPR and thoracotomy" showed no correlation with OFI. This could be
explained by inevitable mortality, thus presenting with a smaller time
frame for mistakes to be made.

AFs
generally exhibited a high specificity relative to their sensitivity
which may be attributed to the outcome being rather uncommon, whilst
every individual AF misses the majority of OFIs. This could suggest that
the AFs are too rigid in their criteria. On the contrary, a too
inclusive criteria might flag several patients, without an OFI [@willis2008interpreting].

@@

AFs, although widely used, there is lack
of consensus on what they are, what purpose they serve and how they
shall be used [@stelfox2010quality; @stelfox2011evidence;
@berg2022perceived]. Further, AFs are static in their criteria, contrary to health care in general.
Consequently, the utility of AFs may diminish with emerging challenges
or upon resolution of the quality gaps they were designed for. This
highlights the need for consistent and evidence-based review to
determine current usefulness [@santana2014development]. However, such
review process might present with costs greater than the potential
benefit from AFs.

Due to the
study being conducted in a single, trauma level I centre in Stockholm it
is unrealistic to assume applicability of these findings at other
hospitals facing unique challenges [@world2009guidelines]. It is previously known that
low and middle income countries seem to benefit from certain AFs that
high income countries do not and vice versa [@berg2022perceived].

@@

Given the expanding field of artificial intelligence and machine
learning, newer studies should focus on using these tools to screen for
patient cases at risk of an OFI. This will diminish misclassification
rates and ensure sustained high performance, attributable
to an algorithm's capability to adapt and learn from the prevailing
context of trauma care. This approach is contemporary but not
unexplored. A recent preprint by Attergrim et al.
[@attergrim2023predicting] explored the idea of machine learning models
as a screening tool for potential OFIs. They found that machine learning
models outperformed the AFs currently used at Karolinska University
Hospital in predicting OFIs.

@@

# Conclusion

AFs are widely used as a tool in QI programmes and
should reflect the current gap in trauma care quality. Strong evidence
supporting their efficacy in quality improvement is scarce. In this
study, no audit filter seemed to show promising performance in
predicting OFIs with AUC ranging from 0.5 to 0.63. These results
highlight the importance of continually evaluating the relevance of AFs.
However, this process takes time and resources and using an alternative
to AFs is recommended.

# References
